Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2017-05-12 pkg/glmertree/paper/treatment_interactions.Rnw
Minor paper revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1328 255f4249-2263-4417-b1ea-8498779e27a7

On average, LMM trees showed somewhat higher accuracy: the average correlation between true and predicted treatment-effect differences was .54 (SD = .40) for LMM trees and .51 (SD = .43) for LMMs. The effects of the most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlation_pw}. As Figure~\ref{fig:xyplot_correlation_pw} indicates, LMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas LMMs show highest predictive accuracy in datasets with purely continuous interactions. LMM trees perform poorly only when interactions are purely linear, whereas LMMs perform poorly when interactions are not purely continuous. Strikingly, Figure~\ref{fig:xyplot_correlations_pw} suggests that LMMs perform somewhat more accurately in the presence of purely piecewise interactions than in the presence of partly continuous interactions, but only with larger sample sizes and a smaller number of potential moderator variables.

Performance of both LMM trees and LMMS improves with increasing sample size. Furthermore, performance of LMM trees is not affected by the number of covariates, whereas the predictive accuracy of LMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous. This indicates that LMM trees are especially useful for exploratory purposes, where there are many potential moderator variables.

Finally, it should be noted that in terms of interpretability, LMM trees often provide simpler models: The LMMs included 12.30 significant terms on average, whereas LMM trees had 3.38 inner nodes on average, requiring only about 3-4 variables to be evaluated for making predictions.

@@

GLMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions, outperforming GLM trees without random effects. In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERTs, predicting treatment-effect differences in test data with .94 accuracy, on average. In datasets without treatment-subgroup effects, GLMM trees showed a Type-I error rate 4\$, compared with a Type-I error rate of 33\% for GLM trees.

@@

Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy, on average. GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise. GLMM trees performed better in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited for exploration that GLMMs. Also, GLMM trees may be easier to interpret and apply: The number of terms in a GLMM increases quadratically with the number of predictors, yielding complex models. GLMM trees in our simulations had a maximum tree depth of three, requiring evaluation of at most three variables to make a prediction for new observations.

In the Application, the GLMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the other two methods (GLM tree and a GLMM with pre-specified interactions). The GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.

>>>>

On average, LMM trees showed somewhat higher accuracy: the average correlation between true and predicted treatment-effect differences was .54 (SD = .40) for LMM trees and .51 (SD = .43) for LMMs. The effects of the most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlation_pw}. As Figure~\ref{fig:xyplot_correlation_pw} indicates, LMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas LMMs show highest predictive accuracy in datasets with purely continuous interactions. LMM trees perform poorly only when interactions are purely linear, whereas LMMs perform poorly when interactions are not purely continuous. Strikingly, Figure~\ref{fig:xyplot_correlation_pw} suggests that LMMs perform somewhat more accurately in the presence of purely piecewise interactions than in the presence of partly continuous interactions, but only with larger sample sizes and a smaller number of potential moderator variables.

Performance of both LMM trees and LMMS improves with increasing sample size. Furthermore, performance of LMM trees is not affected by the number of covariates, whereas the predictive accuracy of LMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous. This indicates that LMM trees are especially useful for exploratory purposes, where there are many potential moderator variables. In addition, LMM trees may often provide simpler models: Whereas the LMMs included 12.30 significant terms on average, LMM trees had 3.38 inner nodes on average, requiring only about 3-4 variables to be evaluated for making predictions.

@@

GLMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions, outperforming GLM trees without random effects. In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERTs, predicting treatment-effect differences in test data with .94 accuracy, on average. In datasets without treatment-subgroup effects, GLMM trees showed a Type-I error rate of 4\%, compared with a Type-I error rate of 33\% for GLM trees.

@@

Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy on average. GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise. GLMM trees performed better in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited for exploration than GLMMs.

In the Application, we obtained similar results: the GLMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the other two methods (GLM tree and a GLMM with pre-specified interactions). The GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.
